---
title: "A quick start guide to the standR package"
author: "Ning Liu, Melissa Davis"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    number_sections: true
    theme: cosmo
    highlight: tango
    code_folding: show
vignette: >
  %\VignetteIndexEntry{hoodscanR_introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<style type="text/css">
  body{
  font-size: 14pt;
}
</style>

```{r message=FALSE, warning=FALSE, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, dpi = 72, fig.retina = 1)
suppressWarnings(library(ggplot2))


```

# Installation

```{r eval=FALSE}
if (!require("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}

BiocManager::install("hoodscanR")
```

The development version of `standR` can be installed from GitHub:

```{r eval=FALSE}
devtools::install_github("DavisLaboratory/hoodscanR")
```

# Quick start

```{r}
library(hoodscanR)
library(SpatialExperiment)
library(scico)
```

# Data exploration

```{r}
data("spe_test")

spe <- readHoodData(spe, anno_col = "celltypes")
```

```{r}
spe
```

```{r}
colData(spe)
```

We can have a look at the tissue and cell positions by using the function `plotTissue`.

The test data is relatively sparse with low-level cell type annotations.

```{r}
col.pal <- c("red3", "royalblue", "gold", "cyan2", "purple3", "darkgreen")

plotTissue(spe, color = cell_annotation, size = 1.5, alpha = 0.8) +
  scale_color_manual(values = col.pal)
```

# Neighborhoods scanning

In order to perform neighborhood scanning, we need to firstly identify k (in this example, k = 100) nearest cells for each cells.

```{r}
fnc <- findNearCells(spe, k = 100)
```

The output of `findNearCells` function includes two matrix, an annotation matrix and a distance matrix.

```{r}
lapply(fnc, function(x) x[1:10, 1:5])
```

We can then perform neighborhood analysis using the function `scanHoods`.

```{r}
pm <- scanHoods(fnc$distance)
```

The resulting matrix is the probability of each cell associating with their 100 nearest cells.

```{r}
pm[1:10, 1:5]
```

We can then merge the probabilities by the cell types of the 100 nearest cells.

```{r}
hoods <- mergeByGroup(pm, fnc$cells)
```

```{r}
hoods[1:10, ]
```

# Neighborhoods analysis

We plot randomly plot 10 cells to see the output of neighborhood scanning using `plotHoodMat`.

```{r}
plotHoodMat(hoods, n = 10, hm_height = 5)
```

Or to check the cells-of-interest with the parameter `targetCells` within the function

```{r}
plotHoodMat(hoods, targetCells = c("Lung9_Rep1_5_1975", "Lung9_Rep1_5_2712"), hm_height = 3)
```

We can then merge the neighborhood results with the SpatialExperiment object using `mergeHoodSpe` so that we can conduct more neighborhood-related analysis.

```{r}
spe <- mergeHoodSpe(spe, hoods)
```

To summarise our neighborhood results, we can use `calcMetrics` to calculate entropy and perplexity of the probability matrix so that we can have a summarisatino of the neighborhood distribution across the tissue slide, i.e. where neighborhood is more distinct and where is more mixed.

```{r}
spe <- calcMetrics(spe, pm_cols = colnames(hoods))
```

We then again use `plotTissue` to plot out the entropy or perplexity.

```{r}
plotTissue(spe, size = 1.5, color = entropy) +
  scale_color_scico(palette = "tokyo")
```

```{r}
plotTissue(spe, size = 1.5, color = perplexity) +
  scale_color_scico(palette = "tokyo")
```

We can perform neighborhood colocalization analysis using `plotColocal`. Here we can see in the test data, endothelial cells and stromal cells are more likely to colocalize, epithelial cells and dividing cells are more likely to colocalize.

```{r}
plotColocal(spe, pm_cols = colnames(hoods))
```

We can cluster the cells by their neighborhood probability distribution using `clustByHood`, it is based on the k-means algorithm and here we set k to 10.

```{r}
spe <- clustByHood(spe, pm_cols = colnames(hoods), k = 10)
```

We can see what are the neighborhood distributions look like in each cluster using `plotProbDist`.

```{r}
plotProbDist(spe, pm_cols = colnames(hoods), by_cluster = TRUE, plot_all = TRUE, show_clusters = as.character(seq(10)))
```

We can plot the clusters on the tissue slide, agian using `plotTissue`.

```{r}
plotTissue(spe, color = clusters)
```

# Session info

```{r}
sessionInfo()
```


